---
title: "Predicting Airbnb Listing Prices in Rome"
output: html_document
date: "2026-07-10"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE
)

# Data
airbnb <- read.csv("../data/listings.csv")

# Data handling
library(tidyverse)     

# Modelling & inference
library(car)            
library(lmtest)        
library(sandwich)       
library(MASS)           
library(broom)          

# Tables & figures
library(knitr)          
library(kableExtra)     
library(stargazer)      
library(modelsummary)   
library(corrplot)       

# Load datasets once
data("affairs")
data("castle")
```

## 1. Introduction

#### 1.1 Business Scenario

Airbnb aims to help hosts determine an optimal nightly price for their
properties. Setting a price too high may reduce booking demand, while
pricing too low can result in lost revenue. To support data-driven
pricing decisions, this project develops a predictive model that
estimates the nightly price of Airbnb listings based on property
characteristics, location, availability, and customer engagement.

#### 1.2 Research Question

> Which listing characteristics have the greatest influence on Airbnb
> prices in Rome, and how accurately can statistical and machine
> learning models predict the nightly price?

#### 1.3 Objectives

The objectives of this project are to:

-   inspect and clean the Airbnb dataset;

-   perform exploratory data analysis (EDA);

-   engineer meaningful features;

-   develop and compare regression and machine learning models;

-   evaluate model performance; and

-   provide business recommendations.

## 2. Data Description

```{r}
library(tidyverse)

airbnb <- read.csv("../data/listings.csv")
```

#### 2.1 Dataset Description

This study uses the **Inside Airbnb Rome** dataset, which contains
information on **37,228 Airbnb listings** located across Rome, Italy.
The dataset includes variables describing listing characteristics,
geographic location, host information, guest engagement, and pricing.

The response variable for this study is the nightly listing **price
(€)**, while the remaining variables serve as potential predictors in
the pricing models developed throughout this project.

#### 2.2 Dataset Overview

The table below summarizes the main characteristics of the dataset.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

overview <- data.frame(
  Characteristic = c(
    "Number of observations",
    "Number of variables",
    "Target variable",
    "Neighbourhoods",
    "Room types",
    "Variables with missing values"
  ),
  Value = c(
    nrow(airbnb),
    ncol(airbnb),
    "Price (€)",
    length(unique(airbnb$neighbourhood)),
    length(unique(airbnb$room_type)),
    sum(colSums(is.na(airbnb)) > 0)
  )
)

kable(
  overview,
  align = c("l","c")
) %>%
  kable_styling(
    bootstrap_options = c("striped","hover"),
    full_width = FALSE,
    position = "center"
  )
```

**Table 1.1 Dataset Overview** The dataset contains a large and diverse
sample of Airbnb listings in Rome, making it suitable for developing
predictive models and identifying the main determinants of listing
prices.

#### 2.3 Variable Dictionary

Table 2 describes the variables included in the dataset.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
variable_dictionary <- data.frame(
  Variable = c(
    "id",
    "name",
    "host_id",
    "host_profile_id",
    "host_name",
    "neighbourhood_group",
    "neighbourhood",
    "latitude",
    "longitude",
    "room_type",
    "price",
    "minimum_nights",
    "number_of_reviews",
    "last_review",
    "reviews_per_month",
    "calculated_host_listings_count",
    "availability_365",
    "number_of_reviews_ltm",
    "license"
  ),
  Description = c(
      "Unique listing identifier",
    "Listing title provided by the host",
    "Unique host identifier",
    "Host profile identifier",
    "Host name",
    "Broad geographical area (if available)",
    "Neighbourhood where the listing is located",
    "Latitude coordinate",
    "Longitude coordinate",
    "Type of accommodation (Entire home, Private room, Shared room, etc.)",
    "Nightly listing price (EUR)",
    "Minimum number of nights required for booking",
    "Total number of reviews received",
    "Date of the most recent review",
    "Average number of reviews per month",
    "Number of active listings managed by the host",
    "Number of days the listing is available during the year",
    "Number of reviews received in the last 12 months",
    "Official registration/license number"
  )
)

kable(
  variable_dictionary
) %>%
  kable_styling(
    bootstrap_options = c("striped","hover"),
    full_width = FALSE,
    position = "center"
  )
```

**Table 2 Variable Dictionary**

#### 2.4 Variables Used in the Analysis

Not all variables contained in the dataset are suitable for predictive
modeling. Variables such as identifiers and descriptive text do not
contribute meaningful predictive information and are therefore excluded
from the analysis.

The variables used for model development are summarized below.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
analysis_variables <- data.frame(
  Variable = c(
    "price",
    "room_type",
    "neighbourhood",
    "latitude",
    "longitude",
    "minimum_nights",
    "number_of_reviews",
    "reviews_per_month",
    "availability_365",
    "calculated_host_listings_count"
  ),
  Role = c(
    "Target variable",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
   "Predictor"
  )
)
kable(
  analysis_variables
) %>%
  kable_styling(
    bootstrap_options = c("striped","hover"),
    full_width = FALSE,
    position = "center"
  )
```

**Table 3 Variables Used** Variables including **id**, **host_id**,
**host_profile_id**, **name**, **host_name**, **license**, and
**last_review** were excluded from the predictive models because serve
solely as identifiers or descriptive text and do not contribute
meaningful predictive information.The remaining variables capture
important characteristics of the listings, their locations, host
activity, and guest engagement, making them appropriate predictors of
nightly listing prices.

## 3. Data Cleaning

Before building predictive models, the dataset was inspected for
potential quality issues that could affect the analysis. The following
checks were performed: -Missing values -Duplicate listings -Variable
data types -Invalid or unrealistic prices -Potential outliers

#### 3.1 Missing Values

The dataset was examined for several potential data quality issues
before analysis. Specifically, missing values, duplicate observations,
variable data types, invalid prices, and potential outliers were
investigated to ensure the dataset was suitable for predictive modeling.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)

# Create missing values table
missing_table <- data.frame(
  Variable = names(airbnb),
  Missing_Values = colSums(is.na(airbnb)),
  Missing_Percent = round(colMeans(is.na(airbnb)) * 100, 2)
)

# Keep only variables with missing values
missing_table <- missing_table %>%
  filter(Missing_Values > 0)

# Rename columns for display
colnames(missing_table) <- c(
  "Variable",
  "Missing Values",
  "Missing (%)"
)

# Display table
kable(
  missing_table,
  align = c("l", "r", "r")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    position = "center"
  )
```

**Table 4. Missing Values by Variable"**

The variable neighbourhood_group contains only missing values and
therefore provides no useful information for the analysis. It will be
removed from the dataset. The target variable price has approximately
7.4% missing values. Since price is the response variable, observations
with missing prices cannot be used to train a predictive model and will
therefore be excluded. The variable reviews_per_month contains around
12% missing values. These missing values are likely associated with
listings that have received no reviews rather than representing data
collection errors. This variable will be investigated further during
exploratory data analysis before deciding whether to impute missing
values or replace them with zero. The remaining variables contain very
few missing values (\<1%) and are unlikely to substantially affect the
analysis.

#### 3.3 Invalid Prices

Summary statistics for the target variable are presented in Table 5.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

price_summary <- data.frame(
  Statistic = c(
    "Minimum",
    "1st Quartile",
    "Median",
    "Mean",
    "3rd Quartile",
    "Maximum"
  ),
  `Price (€)` = c(
    "3",
    "119",
    "162",
    "221.9",
    "234",
    "13,729"
  )
)

kable(
  price_summary,
  align = c("l", "r")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    position = "center"
  )
```

**Table 5 Summary Statistics for Price**

#### 3.4 Initial observations

Overall, the dataset is of high quality and requires only limited
preprocessing. The primary data quality issues are missing values in
price and reviews_per_month, together with the completely empty
neighbourhood_group variable. No duplicate observations or invalid
prices were identified. These observations will be examined further
during exploratory data analysis to determine whether they represent
legitimate luxury properties or influential outliers that require
additional treatment.

```{r}
boxplot(
  airbnb$price,
  ylab = "Price (€)"
)
Q1 <- quantile(airbnb$price, 0.25, na.rm = TRUE)

Q3 <- quantile(airbnb$price, 0.75, na.rm = TRUE)

IQR_value <- IQR(airbnb$price, na.rm = TRUE)

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

```

**Figure 1. Distribution of Airbnb Listing Prices**

Airbnb prices are typically right-skewed, with a small number of luxury
properties creating extreme values. Outliers will be investigated
further before deciding whether to remove, winsorize, or retain them, as
they may represent legitimate high-end listings.

Following the data cleaning process, the dataset is suitable for
exploratory data analysis and predictive modeling. Missing values have
been assessed, duplicate observations were not identified, invalid
prices were excluded where necessary, and potential outliers have been
documented for further investigation.

## 5. Exploratory Data Analysis

Before building predictive models, it is important to understand the
relationships between the response variable and the explanatory
variables. This section explores the distribution of Airbnb prices and
examines how listing characteristics, location, and guest engagement
relate to nightly prices.

#### 5.1 Distribution of Listing Prices

```{r, fig.align='center', fig.width=7, fig.height=5}

library(ggplot2)

ggplot(airbnb, aes(price)) +
  geom_histogram(
    bins = 40,
    fill = "#C71585",
    color = "white"
  ) +
  labs(
    x = "Nightly Price (€)",
    y = "Number of Listings"
  ) +
  theme_minimal()
```

**Figure 2.Distribution of Airbnb Listing Prices**

The histogram shows that Airbnb listing prices in Rome are highly
right-skewed. Most listings are concentrated at relatively low nightly
prices (below approximately €300), while a small number of luxury
properties have exceptionally high prices. These extreme values increase
the mean price and suggest the presence of outliers. Consequently, a log
transformation of the target variable may improve the performance of
predictive models.

#### 5.2 Room Type

```{r, fig.align='center', fig.width=7, fig.height=5}
ggplot(airbnb,
       aes(room_type, price, fill = room_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  labs(
    title = "Listing Price by Room Type",
    x = "Room Type",
    y = "Nightly Price (€)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Figure 3.Listing Price by Room Type**

Room type appears to be one of the strongest determinants of Airbnb
prices. Entire homes and apartments generally command the highest
nightly prices and exhibit the greatest price variability. Private rooms
are typically less expensive, while shared rooms represent the
lowest-priced accommodation type. The large spread within each category
suggests that additional factors, such as location and property
characteristics, also influence listing prices.

#### 5.3 Price by Neighbourhood

```{r, fig.align='center', fig.width=8, fig.height=6}

library(dplyr)

avg_price <- airbnb %>%
  group_by(neighbourhood) %>%
  summarise(
    Average_Price = mean(price, na.rm = TRUE)
  ) %>%
  arrange(desc(Average_Price))

ggplot(avg_price,
       aes(reorder(neighbourhood,
                   Average_Price),
           Average_Price)) +
  geom_col(fill = "#C71585") +
  coord_flip() +
  labs(
    x = "",
    y = "Average Price (€)"
  ) +
  theme_minimal()
```

**Figure 4.Average Airbnb Price by Neighbourhood**

Average listing prices differ considerably across Rome’s neighbourhoods.
The historic city centre (I Centro Storico) has the highest average
nightly price, reflecting its popularity among tourists and proximity to
major attractions. In contrast, peripheral neighbourhoods exhibit
substantially lower average prices. These findings indicate that
neighbourhood is likely to be an important predictor in the pricing
model.

#### 5.4 Reviews and Listing Price

```{r, fig.align='center', fig.width=7, fig.height=5}

ggplot(airbnb,
       aes(number_of_reviews, price)) +
  geom_point(alpha = 0.2,
             color = "#C71585") +
  labs(
    title = "Reviews and Listing Price",
    x = "Number of Reviews",
    y = "Nightly Price (€)"
  ) +
  theme_minimal()
```

**Figure 5.Reviews and Listing Price**

The scatterplot suggests that the relationship between the number of
reviews and listing price is relatively weak. Most highly reviewed
listings are concentrated within lower and medium price ranges, while
expensive listings generally receive fewer reviews. This pattern may
reflect higher booking frequencies for more affordable accommodations.
Although review count alone is not a strong predictor of price, it may
still contribute useful information when combined with other variables.

#### 5.5 Availability and Listing Price

```{r, fig.align='center', fig.width=7, fig.height=5}

ggplot(airbnb,
       aes(availability_365,
           price)) +
  geom_point(alpha = 0.2,
             color = "#C71585") +
  labs(
    x = "Availability (Days per Year)",
    y = "Nightly Price (€)"
  ) +
  theme_minimal()
```

**Figure 6.Availability and Listing Price**

No clear linear relationship is observed between annual availability and
nightly price. Listings with both high and low availability span a wide
range of prices. This suggests that availability alone is unlikely to
explain price differences, although it may capture aspects of host
behaviour or seasonal demand when considered alongside other predictors.

#### 5.6 Minimum Nights and Listing Price

```{r, fig.align='center', fig.width=7, fig.height=5}

ggplot(airbnb,
       aes(minimum_nights,
           price)) +
  geom_point(alpha = 0.2,
             color = "#C71585") +
  labs(
    x = "Minimum Nights",
    y = "Nightly Price (€)"
  ) +
  theme_minimal()
```

**Figure 7.Minimum Stay Requirement and Listing Price**

Most Airbnb listings require only a small minimum stay, typically fewer
than ten nights. Listings with very high minimum-night requirements are
relatively rare and do not appear to follow a consistent pricing
pattern. Overall, the relationship between minimum stay requirements and
listing price appears weak, suggesting that this variable alone has
limited explanatory power.

#### 5.7 Correlation Analysis

Correlation analysis provides an overview of the relationships among the
numerical variables in the dataset. Strong correlations may indicate
variables that contain similar information, while weak correlations
suggest that each predictor contributes unique information.
Understanding these relationships also helps identify potential
multicollinearity before building predictive models.

```{r, fig.align='center', fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

# Select only numeric variables
numeric_data <- airbnb %>%
  dplyr::select(
    price,
    latitude,
    longitude,
    minimum_nights,
    number_of_reviews,
    reviews_per_month,
    calculated_host_listings_count,
    availability_365,
    number_of_reviews_ltm
  )

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

colnames(cor_matrix) <- c(
  "Price",
  "Latitude",
  "Longitude",
  "Min Nights",
  "Reviews",
  "Reviews/Month",
  "Host Listings",
  "Availability",
  "Reviews (LTM)"
)

rownames(cor_matrix) <- c(
  "Price",
  "Latitude",
  "Longitude",
  "Min Nights",
  "Reviews",
  "Reviews/Month",
  "Host Listings",
  "Availability",
  "Reviews (LTM)"
)
# Convert matrix to long format
cor_df <- as.data.frame(as.table(cor_matrix))

# Plot heatmap
ggplot(cor_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Freq, 2)), size = 3) +
  scale_fill_gradient2(
    low = "#FFFFFF",
    mid = "#FCE4EC",
    high = "#C2185B",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    x = "",
    y = "",
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**Figure 8.Correlation Matrix of Numerical Variables**

The correlation matrix indicates that most numerical variables exhibit
weak pairwise correlations, suggesting that multicollinearity is
unlikely to be a major concern in the predictive models. As expected,
the strongest positive correlation is observed between reviews_per_month
and number_of_reviews_ltm (r = 0.84), since both variables measure
recent guest activity. A moderate positive correlation is also present
between number_of_reviews and number_of_reviews_ltm (r = 0.61), and
between number_of_reviews and reviews_per_month (r = 0.54). In contrast,
the target variable price exhibits only very weak correlations with the
numerical predictors, indicating that listing price is likely influenced
by a combination of numerical and categorical characteristics rather
than by any single numerical variable. These findings support the use of
multivariate regression and machine learning models to capture the
complex relationships underlying Airbnb pricing.

Although the numerical predictors display relatively weak correlations
with price, this does not imply that they lack predictive value. Several
important explanatory variables in this dataset, including room type and
neighbourhood, are categorical and therefore are not represented in the
correlation matrix. Moreover, the observed relationships may be
non-linear, making machine learning approaches particularly suitable for
this prediction task.

#### 5.8 Key Findings

The exploratory data analysis revealed several important characteristics
of the Airbnb dataset. First, listing prices are highly right-skewed,
with a relatively small number of luxury properties creating extreme
values. Applying a logarithmic transformation substantially reduced this
skewness, producing a distribution that is more appropriate for
predictive modeling.

Among the explanatory variables, room type and neighbourhood appear to
have the strongest influence on listing prices. Entire homes and
apartments generally command the highest nightly prices, while shared
rooms are consistently the least expensive. Listings located in Rome’s
historic centre also tend to achieve higher average prices than those in
more peripheral neighbourhoods.

The numerical variables, including review activity, availability, and
minimum stay requirements, exhibit relatively weak individual
relationships with price. Likewise, the correlation analysis indicates
limited multicollinearity among the numerical predictors, although
review-related variables are naturally correlated with one another.
These findings suggest that Airbnb pricing is influenced by the combined
effects of multiple listing characteristics rather than by any single
variable.

Overall, the exploratory analysis indicates that the dataset is suitable
for predictive modeling and provides a strong foundation for the feature
engineering and regression analyses that follow.

## 6. Feature Engineering

Before building predictive models, several new variables were created
from the original dataset. Feature engineering can improve predictive
performance by transforming raw information into variables that better
capture important characteristics of Airbnb listings.

#### 6.1 Log Transformation of Price

The histogram presented earlier showed that listing prices are highly
right-skewed, with a small number of luxury properties generating
extreme values. Such skewness may violate the assumptions of linear
regression and reduce predictive accuracy. A common solution is to apply
a logarithmic transformation to the target variable.

```{r}
airbnb$log_price <- log(airbnb$price)
```

```{r, fig.align='center', fig.width=7, fig.height=5}
ggplot(airbnb, aes(log_price)) +
  geom_histogram(
    bins = 40,
    fill = "#C2185B",
    color = "white"
  ) +
  labs(
    x = "Log(Price)",
    y = "Number of Listings"
  ) +
  theme_minimal()
```

**Figure 9. Distribution of Log-Transformed Listing Prices**

The logarithmic transformation substantially reduces the positive
skewness observed in the original price distribution. After
transformation, the distribution becomes considerably more symmetric and
closely resembles a normal distribution, while the influence of extreme
luxury listings is reduced. This transformation is expected to improve
the performance of linear regression models by better satisfying the
assumptions of normality and constant variance.

```{r, fig.align='center', fig.width=7, fig.height=5}
ggplot(airbnb,
       aes(room_type, log_price, fill = room_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  labs(
    x = "Room Type",
    y = "Log(Price)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Figure 10. Log-Transformed Listing Price by Room Type**

The differences in listing prices across room types remain evident after
applying the logarithmic transformation. Entire homes and apartments
continue to exhibit the highest median prices, followed by hotel rooms
and private rooms, while shared rooms remain the least expensive
accommodation type. Compared with the original boxplot, the variability
within each category is noticeably reduced, indicating that the
logarithmic transformation successfully minimizes the influence of
extreme observations while preserving the underlying pricing patterns.

#### 6.2 High Availability

Listings available for more than half of the year may reflect different
pricing strategies than those with limited availability.

A binary variable named `high_availability` was created to identify
listings available for more than 180 days per year. This feature is
intended to capture differences in host pricing strategies and listing
availability and will be included as a predictor in the subsequent
models.

```{r}

airbnb$high_availability <- ifelse(

  airbnb$availability_365 > 180,

  1,

  0

)
```

## 7. Regression Model

#### 7.1 Model Specification

The objective of the regression analysis is to identify the listing
characteristics that influence Airbnb prices in Rome and to evaluate the
predictive performance of multiple linear regression. The dependent
variable is log_price, which represents the natural logarithm of the
nightly listing price. The original price variable exhibited substantial
positive skewness due to a small number of luxury listings. Applying a
logarithmic transformation reduces the influence of these extreme
observations and produces a distribution that better satisfies the
assumptions of linear regression. The explanatory variables were
selected based on economic reasoning and the exploratory data analysis
conducted in the previous sections. confirm the OLS findings.

**Main Explanatory Variables:**

Room Type (room_type) - Room type is expected to be one of the strongest
determinants of Airbnb prices. Entire homes and apartments generally
provide greater privacy and larger living spaces than private or shared
rooms and therefore command higher nightly rates.

Neighbourhood (neighbourhood) - Location is a fundamental determinant of
housing prices. Listings located in central tourist areas are expected
to be more expensive because of greater accessibility and higher demand.

**Control Variables**

Minimum stay requirements(minimum_nights) may reflect different pricing
strategies adopted by hosts. Listings requiring longer stays may either
charge higher prices because they target longer-term guests or lower
prices to encourage extended bookings. Therefore, the expected
relationship is ambiguous.

Review count(number_of_reviews) measures guest engagement and
popularity. Highly reviewed properties may attract more demand and
therefore charge higher prices. However, older listings also tend to
accumulate more reviews over time, making the expected relationship
uncertain.

This variable captures recent booking activity(reviews_per_month).
Listings receiving reviews more frequently may reflect higher demand and
stronger market performance, suggesting a positive association with
price.

Availability (availability_365) measures the number of days a listing is
available throughout the year. Properties with limited availability may
reflect stronger demand, while highly available properties may indicate
lower occupancy rates. Consequently, the expected relationship is
ambiguous.

Hosts managing multiple properties may adopt professional pricing
strategies that differ from individual
hosts(calculated_host_listings_count) . The direction of this
relationship is determined empirically.

| Variable          | Expected Effect on Price            |
|-------------------|-------------------------------------|
| Room Type         | Positive (+)                        |
| Neighbourhood     | Positive/Negative (depends on area) |
| Minimum Nights    | Ambiguous (+/-)                     |
| Number of Reviews | Positive (+)                        |
| Reviews per Month | Positive (+)                        |
| Availability      | Ambiguous (+/-)                     |
| Host Listings     | Ambiguous (+/-)                     |

**Table 6 Expected Coefficient Signs**

**Baseline Model**

The baseline regression model examines the relationship between room type and the logarithm of the nightly Airbnb listing price:

$$
\log(\text{Price}_i)=
\beta_0+
\beta_1\text{RoomType}_i+
u_i
$$

where u_i represents unobserved factors affecting the listing price.

**Extended Model**

To account for additional listing characteristics and guest engagement, the following specification is estimated:

$$
\log(\text{Price}_i)=
\beta_0+
\beta_1\text{RoomType}_i+
\beta_2\text{MinimumNights}_i+
\beta_3\text{Reviews}_i+
\beta_4\text{ReviewsPerMonth}_i+
\beta_5\text{Availability}_i+
u_i
$$
This specification evaluates whether booking characteristics and customer engagement improve the explanation of Airbnb listing prices beyond accommodation type alone.

**Full Model**

The final specification additionally controls for neighbourhood and host characteristics:
$$
\log(\text{Price}_i)=
\beta_0+
\beta_1\text{RoomType}_i+
\beta_2\text{Neighbourhood}_i+
\beta_3\text{MinimumNights}_i+
\beta_4\text{Reviews}_i+
\beta_5\text{ReviewsPerMonth}_i+
\beta_6\text{Availability}_i+
\beta_7\text{HostListings}_i+
u_i
$$

This model provides the most comprehensive specification by incorporating property characteristics, geographic location, booking requirements, guest engagement, and host activity. Comparing the three models allows us to evaluate how the inclusion of additional predictors improves the explanatory power of the regression model.


### 7.2. Regression Analysis

To examine the factors influencing Airbnb listing prices in Rome, a
series of Ordinary Least Squares (OLS) regression models are estimated.
The dependent variable in all specifications is the log-transformed
nightly listing price (log_price), while listing characteristics,
location, availability, and guest engagement serve as the primary
explanatory variables.

The analysis begins with a baseline model that includes only room type
as the main predictor. Additional specifications progressively
incorporate neighbourhood, booking characteristics, host information,
and guest engagement variables to evaluate whether the explanatory power
of the model improves as more relevant predictors are included. This
stepwise approach allows the contribution of each group of variables to
be assessed while examining the robustness of the estimated
relationships..

```{r}

# Model 1: Baseline model
model1 <- lm(
  log_price ~ room_type,
  data = airbnb
)

# Model 2: Add booking and engagement variables
model2 <- lm(
  log_price ~
    room_type +
    minimum_nights +
    number_of_reviews +
    reviews_per_month +
    availability_365,
  data = airbnb
)
# Model 3: Full model
model3 <- lm(
  log_price ~
    room_type +
    neighbourhood +
    minimum_nights +
    number_of_reviews +
    reviews_per_month +
    availability_365 +
    calculated_host_listings_count,
  data = airbnb
)
```

```{r regression-table,results='asis'}
library(stargazer)

stargazer(

  model1,

  model2,

  model3,

  type = "html",

  title = "Comparison of Multiple Linear Regression Models",

  column.labels = c("Model 1", "Model 2", "Model 3"),

  dep.var.labels = "Log Listing Price",

  digits = 3,

  out.header = FALSE

)
```

**Table 7 Comparison of OLS Regression Models**

**M1 = Baseline model, M2 = Extended model, M3 = Full model**

Table 7 presents the results of three Ordinary Least Squares (OLS) regression models estimating the determinants of Airbnb listing prices in Rome. The dependent variable in all specifications is the logarithm of the nightly listing price.

Model 1 includes only room type and serves as the baseline specification. The results indicate that accommodation type is a strong predictor of listing prices. Relative to entire homes and apartments (the reference category), hotel rooms, private rooms, and shared rooms are all associated with significantly lower prices. Shared rooms exhibit the largest negative coefficient, indicating that they are the least expensive accommodation type.

Model 2 extends the analysis by incorporating booking characteristics and guest engagement variables. Minimum nights and reviews per month are negatively associated with listing prices, suggesting that listings requiring longer minimum stays or receiving reviews more frequently tend to charge slightly lower prices after controlling for room type. In contrast, availability throughout the year has a small but positive relationship with price, while the total number of reviews is not statistically significant.

Model 3 provides the most comprehensive specification by additionally controlling for neighbourhood and host characteristics. The inclusion of neighbourhood substantially improves the explanatory power of the model, with the adjusted R^2 increasing from 0.202 in Model 2 to 0.354. This finding highlights the importance of location in determining Airbnb prices. Nearly all neighbourhood coefficients are negative relative to the reference neighbourhood (Centro Storico), indicating that listings outside the historic centre generally command lower nightly prices.

The coefficient for minimum nights remains negative and statistically significant across both extended specifications, while availability continues to exhibit a small positive association with price. Reviews per month remains negatively associated with price, whereas the total number of reviews becomes weakly negative but remains economically small. Finally, the number of listings managed by a host has a positive and statistically significant coefficient, suggesting that professional hosts operating multiple properties tend to charge slightly higher prices.

Overall, the regression results demonstrate that room type and neighbourhood are the primary determinants of Airbnb prices in Rome, while booking characteristics and host information provide additional explanatory power. The progressive increase in the adjusted R^2 from 0.124 in the baseline model to 0.354 in the full specification indicates that the inclusion of location and host characteristics substantially improves model performance. Although the regression explains approximately one-third of the variation in listing prices, a considerable proportion of price differences remains unexplained, suggesting that additional factors—such as property size, amenities, interior quality, and seasonal demand—may also influence Airbnb pricing but are not captured in the available dataset.

```{r residual-diagnostics,fig.width=8, fig.height=8}
# Show the four standard regression diagnostic plots for Model 3 in a 2x2 grid
par(mfrow = c(2, 2))   # arrange the next plots in 2 rows x 2 columns
plot(model3)           # base R automatically produces the 4 lm diagnostic plots
```

**Figure 11 Residual Diagnostics**

Figure 11 presents the standard diagnostic plots for the final regression model (Model 3). These plots assess whether the assumptions of linear regression are reasonably satisfied.

The Residuals vs Fitted plot shows that the residuals are generally centered around zero, indicating that the linear specification is appropriate. However, the spread of the residuals increases slightly for larger fitted values, suggesting the presence of heteroskedasticity, where the variance of the errors is not constant.

The Normal Q-Q plot indicates that the residuals deviate from the reference line in the upper tail. This suggests that the residuals are not perfectly normally distributed, primarily because the dataset contains a number of very expensive Airbnb listings that create extreme observations. Such departures from normality are common in pricing data.

The Scale-Location plot also suggests mild heteroskedasticity. The red smoothing line is not perfectly horizontal, indicating that the variability of the residuals changes across the range of fitted values. This pattern is consistent with the increasing dispersion observed in the Residuals vs Fitted plot.

Finally, the Residuals vs Leverage plot identifies several observations with relatively high leverage. However, none appear to exceed Cook’s distance thresholds by a substantial margin, indicating that no single observation has an excessive influence on the estimated regression coefficients.

Overall, the diagnostic plots indicate that the model provides a reasonable fit to the data, although the assumptions of constant error variance and normality are only partially satisfied. These violations are expected when modelling Airbnb prices because listing prices are naturally right-skewed and include a small number of luxury properties. The earlier logarithmic transformation substantially improved the distribution of the response variable and reduced the influence of extreme values, making the regression results more reliable.

## 8. Machine Learning

While multiple linear regression provides an interpretable baseline model, machine learning algorithms may achieve higher predictive accuracy by capturing complex nonlinear relationships and interactions among the predictors. This section compares several predictive models to evaluate their ability to estimate Airbnb listing prices in Rome.

The cleaned dataset is randomly divided into a training set (80%) and a testing set (20%). The models are fitted using the training data, while the testing data are used to evaluate out-of-sample prediction performance. This approach helps assess how well each model generalizes to unseen observations and reduces the risk of overfitting.

Three predictive models are considered:

* Multiple Linear Regression
* Regression Tree
* Random Forest

The performance of each model is compared using the Root Mean Squared Error (RMSE), where lower values indicate better predictive accuracy.

```{r}
# Keep only the variables used in the models
airbnb_model <- airbnb[, c(
  "log_price",
  "room_type",
  "neighbourhood",
  "minimum_nights",
  "number_of_reviews",
  "reviews_per_month",
  "availability_365",
  "calculated_host_listings_count"
)]

# Remove rows with missing values
airbnb_model <- na.omit(airbnb_model)

# Train-test split
set.seed(123)

train_index <- sample(
  1:nrow(airbnb_model),
  0.8 * nrow(airbnb_model)
)

train <- airbnb_model[train_index, ]
test  <- airbnb_model[-train_index, ]
```

```{r}
lm_model <- lm(
  log_price ~ room_type +
    neighbourhood +
    minimum_nights +
    number_of_reviews +
    reviews_per_month +
    availability_365 +
    calculated_host_listings_count,
  data = train
)

lm_pred <- predict(lm_model, newdata = test)

lm_rmse <- sqrt(mean((test$log_price - lm_pred)^2, na.rm = TRUE))
```

```{r}
library(rpart)

tree_model <- rpart(
  log_price ~ room_type +
    neighbourhood +
    minimum_nights +
    number_of_reviews +
    reviews_per_month +
    availability_365 +
    calculated_host_listings_count,
  data = train,
  method = "anova"
)

tree_pred <- predict(tree_model, newdata = test)

tree_rmse <- sqrt(mean((test$log_price - tree_pred)^2, na.rm = TRUE))
```


```{r}
library(randomForest)

set.seed(123)

rf_model <- randomForest(
  log_price ~ room_type +
    neighbourhood +
    minimum_nights +
    number_of_reviews +
    reviews_per_month +
    availability_365 +
    calculated_host_listings_count,
  data = train,
  ntree = 300
)

rf_pred <- predict(rf_model, newdata = test)

rf_rmse <- sqrt(mean((test$log_price - rf_pred)^2, na.rm = TRUE))
```

```{r}
results <- data.frame(
  Model = c(
    "Multiple Linear Regression",
    "Regression Tree",
    "Random Forest"
  ),
  RMSE = c(
    lm_rmse,
    tree_rmse,
    rf_rmse
  )
)

knitr::kable(
  results,
  digits = 3
)
```
**Table 12.Comparison of Prediction Accuracy **

Table 12 compares the predictive performance of the three models using the Root Mean Squared Error (RMSE), where lower values indicate more accurate predictions.

The Multiple Linear Regression model achieved an RMSE of 0.481, providing a strong and interpretable baseline for predicting Airbnb listing prices. Although the regression model captures the main linear relationships between the explanatory variables and listing prices, its predictive accuracy is limited by its inability to model complex nonlinear patterns.

The Regression Tree produced an RMSE of 0.502, which is slightly higher than that of the linear regression model. This indicates that the regression tree provides less accurate predictions for this dataset. While regression trees are capable of modelling nonlinear relationships, they are also prone to producing simpler decision rules that may not fully capture the complexity of Airbnb pricing.

The Random Forest model achieved the lowest RMSE (0.465) among all three models, making it the best-performing model in this analysis. By combining the predictions of many decision trees, the random forest is able to model complex interactions and nonlinear relationships while reducing overfitting. The improvement in predictive accuracy suggests that Airbnb listing prices are influenced by relationships that cannot be fully captured by a traditional linear regression model.

Overall, the comparison demonstrates that machine learning techniques can improve the prediction of Airbnb listing prices. Although the multiple linear regression model remains valuable because of its interpretability, the random forest provides the highest predictive accuracy and is therefore the preferred model for forecasting Airbnb prices in Rome.



```{r, fig.align='center', fig.width=8, fig.height=6}
library(randomForest)
varImpPlot(
  rf_model,
  main = "Random Forest Feature Importance",
  col = "#C2185B"
)
```

**Figure 12. Variable Importance from the Random Forest Model**

The variable importance plot identifies the predictors that contribute most to the Random Forest model. Among the variables included in the analysis, neighbourhood is the most important predictor of Airbnb listing prices, followed by room type. This indicates that location and accommodation type are the primary determinants of pricing in the Rome Airbnb market.

Review activity, represented by reviews per month and number of reviews, also contributes to the prediction, although to a lesser extent. Variables related to booking characteristics, such as availability, minimum nights, and the number of listings managed by the host, have relatively smaller effects on predictive performance.

These results are consistent with the findings from the exploratory data analysis, where central neighbourhoods and entire homes were associated with higher listing prices.


## 9. Business Recommendations

The findings of this analysis provide several practical recommendations for Airbnb hosts operating in Rome. By understanding the factors that influence listing prices, hosts can make more informed pricing and marketing decisions to improve both occupancy rates and revenue.

**Differentiate pricing by room type** : The regression analysis shows that room type is one of the strongest determinants of listing price. Entire homes and apartments consistently command higher nightly rates than private or shared rooms. Hosts should therefore benchmark their prices against listings offering the same type of accommodation rather than comparing themselves to the entire market.

**Consider location when setting prices ** : The neighbourhood variable was among the most important predictors in both the regression and Random Forest models. Properties located in more desirable or central areas are able to charge substantial price premiums. Hosts should regularly compare their prices with similar listings within their own neighbourhood instead of applying a uniform pricing strategy across different locations.

**Encourage guest engagement and reviews** : Listings with greater guest activity, measured by the number of reviews and reviews per month, tend to achieve higher prices. Maintaining high service quality and encouraging satisfied guests to leave reviews may improve both listing visibility and pricing potential over time.

**Avoid unnecessarily restrictive booking policies ** : The analysis suggests that higher minimum-night requirements are associated with slightly lower listing prices. Setting excessively long minimum stays may reduce demand by limiting booking flexibility. Hosts should carefully balance occupancy goals with pricing strategies.

**Use data-driven pricing strategies ** : The superior performance of the Random Forest model demonstrates that Airbnb prices depend on complex interactions between multiple listing characteristics rather than on a single factor. Hosts may therefore benefit from dynamic pricing tools that simultaneously consider location, accommodation type, availability, and guest engagement when determining nightly prices.


## 10. Conclusion

This study is subject to several limitations. The analysis uses cross-sectional data for Airbnb listings in Rome and therefore does not capture seasonal price fluctuations or changes over time. Additionally, some potentially important determinants of price, such as property amenities, photographs, and textual listing descriptions, were not included in the dataset. Future research could incorporate these variables and explore more advanced machine learning methods to further improve predictive accuracy.

This project examined the factors influencing Airbnb listing prices in Rome using a combination of exploratory data analysis, multiple linear regression, and machine learning techniques. The analysis demonstrated that Airbnb pricing is driven by several property characteristics, with location and room type emerging as the most important determinants of nightly prices.

The exploratory data analysis revealed that listing prices are highly right-skewed, motivating the use of a logarithmic transformation of the target variable. The transformed distribution better satisfied the assumptions of linear regression and improved model stability. Correlation analysis showed that no single numerical variable exhibited a strong linear relationship with price, suggesting that pricing depends on the combined effects of multiple listing characteristics.

Three predictive models were compared: Multiple Linear Regression, Regression Tree, and Random Forest. Among these, the Random Forest model achieved the best predictive performance, producing the lowest Root Mean Squared Error (RMSE = 0.465). This indicates that machine learning methods are better able to capture the complex nonlinear relationships that characterize Airbnb pricing. Nevertheless, the multiple linear regression model remained valuable because of its straightforward interpretation and its ability to quantify the effect of individual predictors.

The analysis also identified several practical insights for Airbnb hosts. Entire homes command higher prices than private or shared rooms, listings in central neighbourhoods benefit from significant price premiums, and guest engagement variables such as review activity contribute positively to pricing. These findings suggest that successful pricing strategies should account for both property characteristics and market conditions rather than relying on simple fixed pricing rules.

Overall, this project demonstrates a complete data analytics workflow, including data cleaning, exploratory analysis, feature engineering, statistical modeling, machine learning, model evaluation, and the translation of analytical results into actionable business recommendations. The findings highlight how data-driven approaches can support more accurate pricing decisions and improve revenue optimization in the short-term rental market.







#Bibliography 

https://insideairbnb.com/it/get-the-data/?utm_source=chatgpt.com